---
title: "Tips for downloading data in R"
date: "Feb 12, 2016"
number_sections: yes
---

```{r global_options, include=FALSE}
knitr::opts_chunk$set(warning=FALSE, message=FALSE, eval=FALSE)
```

# Avoid hard-coding directory paths

This is more of a general tip for reproducibility. For example, we can reference directories at the same level as the working directory by getting the parent directory

```{r}
parent <- dirname(getwd())
```

and then using `file.path` to create the file path. Use `file.path` instead of `paste` to write platform indepedent code. Please see the examples below.

# csv files

For csv files, we need to set `mode="wb"` to tell R to write a binary file. Otherwise `download.file` adds an extra space between records.

```{r}
# EIA residential energy consumption survey (RECS) 27 MB
download.file(url="http://www.eia.gov/consumption/residential/data/2009/csv/recs2009_public.csv",
  mode="wb",
  destfile=file.path(parent,'data','recs2009.csv'))
```

# zip files

To download zip files, we don't need any additional arguments to `download.file`

```{r}
# Porto taxi cab data, 520 MB zip file
download.file(url="https://archive.ics.uci.edu/ml/machine-learning-databases/00339/train.csv.zip",
  destfile=file.path(parent,'data','train.zip', sep=""))
```

We'll unzip the file for demonstration, but this isn't necessary, because R's read functions accept zipped files. By default, `unzip` does not remove the original zip file. 

```{r}
unzip("train.zip")
```

# gz files

Downloading gz files is the same as zip files.

```{r}
# weather data, 206 MB gz file
download.file(url="http://www1.ncdc.noaa.gov/pub/data/ghcn/daily/by_year/2010.csv.gz",
  mode="wb",
  destfile=file.path(parent, 'data', '2010.csv.gz'))
```

To unzip gz files, you need the `R.utils` package, though as for zip files, this usually isn't necessary. `unzip` removes the original gz file by default, but you can set `remove=FALSE` to keep the gz file.

```{r}
ifelse(!require(R.utils), install.packages("R.utils"), "already installed")
library(R.utils)
gunzip("2010.csv.gz", remove=FALSE)
```

---

[Brian Segal](https://github.com/bdsegal)
